---
title: "Intro to Transcriptomics"
author: Nicole Black, Wade Boohar, Kayla Xu
date: 07/17/22
updated: 09/26/23
---

***Deliverables***
-Upload this R Notebook to your GitHub and submit the link to your Repo on BB.
-Include any graphs or figures created in this assignment in the folder with your R notebook with descriptive file names.

Since this is an optional partner activity, it is okay if your answers are the same as your partner’s as long as everyone understands it and could explain it in their own words if asked. Each person must individually push their code to Github. *At the top of your R Notebook, write the name of you and your partner(s) as a comment.*

***Complete the following coding activity and answer any following questions as comments in your R Notebook***

In Transcriptomics_Tutorial.Rmd, you learned how to manipulate the SummarizedExperiment data structure, turn it into more readable dataframes, complete basic analysis using those dataframes, and performed a differential expression analysis based on age. Before saving the rna_counts, rna_clinical, and rna_genes however, we masked out any sample that was not a tumor. In this assignment, you will need to reinstantiate the rna_se object and then perform differential expression analysis based on tumor status.

*Pre-Assignment*
Use knitr function to set your working directory to your analysis_data folder in qbio_490_name.
```{r}
normalizePath("/Users/hunter/Desktop/QBIO_490_HunterN/analysis_data") 
```

Load all necessary packages and instantiate the rna_se object
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
```


*1*
Perform all data cleaning and label manipulation steps (hint: applying colnames and rownames to make the dataframes more readable) except for masking out the patient with an NA for age and the "Solid Tissue Normal" sample types.
```{r}
query <- GDCquery(project = "TCGA-BRCA", data.category = "Transcriptome Profiling", data.type = "Gene Expression Quantification")
GDCdownload(query)
se <- GDCprepare(query)
se <- se[, !is.na(colData(se)$age)]
se <- se[, colData(se)$sample_type != "Solid Tissue Normal"]
colnames(se)[1:5] <- c("SampleID", "GeneSymbol", "TranscriptID", "Expression_R1", "Expression_R2")

```


*2*
Determine and define your experimental design for the differential expression analysis. Control for covariance of age, gender, OR vital_status. Manipulate those columns so that they are ready for differential expression analysis (hint: what kind of variables are they? what data type are they by default? do you need to handle unknown values?)
```{r}
library(DESeq2)

counts = SummarizedExperiment(DFrame())
metadata = SummarizedExperiment(GRanges())

dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design = ~ age + gender + vital_status)

dds$age <- as.numeric(dds$age)
dds$vital_status <- as.factor(dds$vital_status)
dds$gender <- as.factor(dds$gender)
dds <- DESeq(dds)
results <- results(dds)
topGenes <- head(results[order(results$padj), ], n = 10)
print(topGenes)

```


*3*
Perform the differential expression analysis, create a volcano plot, and interpret what genes from each of the 6 sections means (assuming thresholds of -log10padj = 2 and log2FoldChange = 1)
```{r}
dds <- DESeq(dds)
results <- results(dds)
threshold_padj <- 2
threshold_log2FoldChange <- 1
significant_genes <- subset(results, padj < 10^(-threshold_padj) & abs(log2FoldChange) > threshold_log2FoldChange)

# Load necessary libraries for plotting
library(ggplot2)

volcano_plot <- ggplot(results, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 10^(-threshold_padj) & abs(log2FoldChange) > threshold_log2FoldChange, "Significant", "Not Significant"))) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") + 
  geom_vline(xintercept = c(-threshold_log2FoldChange, threshold_log2FoldChange), linetype = "dashed", color = "blue") +
  labs(x = "log2(Fold Change)", y = "-log10(padj)") +
  theme_minimal() +
  ggtitle("Volcano Plot")

print(volcano_plot)

```
Cannot get DESeq to download so volcano cannot print. However, I can predict what will plot in each section.

top-right genes: low p-value, most upregulated, high log2fold change
bottom-right genes: high negative log2fold change, downregulated, low p-value
top-left genes: not significant p-values, high log2fold
bottom-left genes: high negative log2fold, not significant p-values
top-middle genes: slightly high log2fold, slightly higher p-values than bottom-middle
bottom-middle genes: lower p-values, lower log2fold

Save the picture of the volcano plot (using either ggsave() or right clicking and manually downloading the image and push this .Rmd and the image to GitHub)
